\section{Introduction}
\label{sec:intro}

A user sends a request: ``Solve this differential equation step by step and return the answer as \LaTeX.'' You have 13 endpoints: seven LLMs at different price points, three agents, two coding assistants, and a script library. Which one do you call?

This is the routing problem. Get it right and you save money while matching quality. Get it wrong and you either pay for an expensive model that adds nothing, or you pick a cheap one that cannot handle the task.

\subsection{Why This Matters}

The conversation about AI cost centers on which model to use: frontier versus open source, GPT-4 versus Claude versus Llama. That is the wrong frame. The real leverage is not in the model. It is in the orchestration.

Organizations will spend billions of dollars on AI tokens annually. Most of that spend flows through routing decisions that are manual, heuristic, or nonexistent. Effective routing (sending each task to the right endpoint at the right cost) could redirect billions of dollars of compute. Not cut it. Use it better.

As AI systems grow more heterogeneous (LLMs, agents, scripts, tool-use systems, workflows), routing becomes critical infrastructure. The difference between organizations that get the best outcome per dollar of compute and organizations that overpay for mediocre results is not which model they picked. It is whether they route intelligently.

\subsection{The Hypothesis}

We approach this with a specific hypothesis: every endpoint (LLM, API, script, agent, workflow) and every task (user prompt, system request, batch job) can be decomposed into measurable properties. If an endpoint's capabilities can be quantified and a task's requirements can be quantified, then the match between them can be scored. If it can be scored, it can be optimized.

This is not obvious. The dominant assumption is that routing requires either human judgment, learned classifiers trained per model pair, or cascading heuristics. We propose that routing is a measurement problem: score every endpoint against every task through a single equation, then pick the highest score.

\subsection{What Existing Routers Do}

Existing routers have made real progress, but they share assumptions that limit them.

FrugalGPT~\citep{chen2023frugalgpt} introduced cascading: try the cheap model, escalate if it is not confident. This cuts costs 50 to 90\% for simple workloads but assumes models are ordered from weak to strong, which breaks when a cheap coding model outperforms an expensive general model on Python tasks. RouteLLM~\citep{ong2024routellm} trains pairwise classifiers on Chatbot Arena data. Strong for binary routing. Extending pairwise classifiers to $N$ endpoints requires $O(N^2)$ classifiers. GraphRouter~\citep{feng2024graphrouter} uses GNN-learned embeddings over task-model graphs. \citet{dekoninck2024cascade} unified routing and cascading. RouterBench~\citep{hu2024routerbench} standardized evaluation across 11 methods.

These systems share one assumption: they route between LLMs only. Most use additive or linear scoring (\citeauthor{dekoninck2024cascade}'s $\tau_i = \hat{q} - \lambda\hat{c}$, BEST-Route's cost-quality selection). The learned-representation methods (GraphRouter, ZeroRouter, EmbedLLM) use dot-product scoring, which avoids the additive trap but still lacks hard constraint enforcement: a model that cannot fit the input can still score high. None routes to non-LLM endpoints. Scripts have no context window. Agents have no token cost. MCP tools support specific operations but not arbitrary prompts. Routing across these types requires constraint enforcement that additive and embedding-based scores do not provide.

\subsection{What S(M,T) Does}

S(M,T) is a single equation that scores any endpoint against any task. Three terms, multiplied:

\begin{equation*}
\smt = \underbrace{\left[\prod_{j} g_j(M,T)\right]}_{\text{Gates}} \cdot \underbrace{\left[\sum_{i} w_i \cdot \phii(M,T)\right]}_{\text{Compatibility}} \cdot \underbrace{e^{-\betavec(T) \cdot \mathbf{H}(M)}}_{\text{Cost penalty}}
\end{equation*}

\textbf{Gates} check hard constraints: can this endpoint handle this task at all? Binary pass/fail. If any gate fails, the score is zero. \textbf{Compatibility} scores how well the endpoint fits the task across 16 dimensions. \textbf{Cost} penalizes expensive endpoints with a task-conditional temperature: a medical query tolerates high cost, a formatting task does not.

The multiplication is the key design choice. These layers are hierarchical, not interchangeable. A failed gate cannot be compensated by high compatibility. Additive scoring allows this. Multiplicative scoring does not.

The techniques are standard. The gate layer is a Product-of-Experts architecture~\citep{hinton2002poe} with structural parallels to hierarchical MoE gating~\citep{jacobs1991adaptive, jordan1994hierarchical}. The compatibility term is an additive utility function from MCDA~\citep{fishburn1967additive}. The cost penalty is a Boltzmann factor from discrete choice theory~\citep{mcfadden1974conditional, train2009discrete}. Online weight learning uses Robbins-Monro stochastic approximation~\citep{robbins1951stochastic}. We do not claim novelty for these components individually. The contribution is the assembled framework and what we found when we tried to implement it.

\subsection{Why This Direction Matters}

If the hypothesis holds (endpoints and tasks are mathematically decomposable), then routing reduces to measurement. Measure the right properties, and the equation does the rest. Any new model, agent, or tool can be onboarded by measuring its properties and feeding them into the same scoring function. No retraining. No per-pair classifiers. One equation for the entire ecosystem.

We tested this. We designed 16 compatibility functions by hand: reasoning depth, code ability, instruction following, cost efficiency, and twelve others. We tried to compute them from 2.76M public benchmark records. Fourteen designs, four research cycles. All fourteen failed.

The reasons are structural, not fixable with more data: benchmarks use incomparable scales, performance does not transfer from evaluation to deployment, and complex properties lose their task-conditional structure when reduced to single numbers. We call this the \textit{measurement gap}.

But the direction was sound. The equation structure (gates $\times$ compatibility $\times$ cost) was right. The problem was how we computed compatibility. When we replaced the hand-designed functions with 16 learned bilinear forms, the same public data that failed hand-designed measurement achieved 83.63\% accuracy on RouterBench and 94.35\% quality retention at 50\% cost on RouteLLM. The model generalizes across both benchmarks without benchmark-specific tuning.

The measurement gap is the real bottleneck. Not compute. Not model size. Close it, and routing becomes a solved infrastructure problem.

\subsection{Contributions}

\begin{enumerate}
    \item \textbf{S(M,T)}, a scoring framework that routes across heterogeneous endpoints (LLMs, agents, scripts, tool-use systems) through a single equation. The multiplicative gate structure enables type-specific constraint enforcement. No existing router handles this mix. The techniques are standard; the application to heterogeneous routing is new.

    \item \textbf{The measurement gap}, a systematic documentation showing 0 of 14 hand-designed scoring functions can be reliably computed from public benchmarks, with five identified structural causes.

    \item \textbf{Learned compatibility functions} that achieve cross-benchmark generalization, with evidence that data quality, not model capacity, is the binding constraint.

    \item \textbf{The hypothesis} that routing is a measurement problem: if endpoints and tasks are mathematically decomposable, a single equation can score any endpoint against any task. The cross-benchmark results are early evidence that this decomposition is achievable.
\end{enumerate}
