\section{Related Work}
\label{sec:related}

\paragraph{Routing Architectures.}
FrugalGPT~\citep{chen2023frugalgpt} cascades from cheap to expensive models. RouteLLM~\citep{ong2024routellm} trains pairwise classifiers for binary routing. RouterBench~\citep{hu2024routerbench} standardized evaluation. Hybrid LLM~\citep{ding2024hybridllm} routes by difficulty. AutoMix~\citep{madaan2023automix} uses self-verification for escalation. \citet{dekoninck2024cascade} unified routing and cascading. BEST-Route~\citep{ding2025bestroute} jointly selects models and response counts. MoMA~\citep{moma2025routing} and DAAO~\citep{daao2025} extend to LLM-backed agents but not to deterministic scripts or non-AI tools. Each contribution is real. The shared limitation: LLM-to-LLM routing with continuous scoring.

S(M,T) differs structurally. The gate layer enforces hard constraints through multiplication~\citep{fishburn1967additive, train2009discrete}: failing any gate produces zero. The framework routes across heterogeneous endpoints (7 LLMs, 3 agents, 1 script, 2 MCP tool-use systems) through one equation. No existing framework handles this mix.

\paragraph{Learned Routing Representations.}
RouterDC~\citep{chen2024routerdc} trains dual contrastive embeddings. EmbedLLM~\citep{zhuang2024embedllm} learns compact model vectors for performance prediction. ICL-Router~\citep{wang2025iclrouter} uses in-context vectors. Prompt-to-Leaderboard~\citep{frick2025p2l} predicts Bradley-Terry coefficients from prompts. ZeroRouter~\citep{yan2026zerorouter} creates model-agnostic latent spaces for zero-shot onboarding.

GraphRouter~\citep{feng2024graphrouter} is the closest architectural comparison. Both systems learn interaction representations between tasks and models rather than hand-designing scoring rules. The differences are structural. GraphRouter's reward labeling uses a weighted additive formula ($\alpha \cdot \text{Performance} - \beta \cdot \text{Cost}$), meaning a model that fails a hard constraint can still receive a positive reward if its performance score is high enough. S(M,T) enforces hard constraints through multiplicative gates: failing any gate produces zero, regardless of compatibility or cost. GraphRouter also assumes all endpoints are LLMs, while S(M,T)'s gate layer encodes type-specific constraints for heterogeneous endpoints. GraphRouter's GNN architecture captures richer relational structure between tasks and models; S(M,T)'s bilinear forms are simpler but integrate into a framework with explicit constraint enforcement and task-conditional cost.

Our BilinearPhiEngine is in this learned-representation family. The architecture (multi-head bilinear forms) is structurally a factorization machine~\citep{rendle2010factorization}, a well-established approach in recommendation systems. What differs is the integration: the learned forms serve as drop-in replacements for the hand-designed phi functions within the S(M,T) framework. The gate layer, cost penalty, and weight learning structure remain unchanged. This modularity is by design: the measurement gap motivates replacing the compatibility functions specifically, while keeping constraint enforcement and cost structure intact.

\paragraph{Online Learning.}
Thompson Sampling for contextual bandits~\citep{agrawal2013thompson} provides the regret framework. BaRP~\citep{barp2025} models routing as multi-objective contextual bandits. OmniRouter~\citep{mei2025omnirouter} uses Lagrangian dual decomposition under budget constraints. We combine Robbins-Monro weight updates~\citep{robbins1951stochastic} with SAC-style entropy decay~\citep{haarnoja2018sac}.
