\documentclass[11pt,a4paper]{article}

% === Typography ===
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{mathptmx}           % Times Roman for text and math
\usepackage[scaled=0.92]{helvet} % Helvetica for sans-serif
\usepackage{courier}             % Courier for monospace
\usepackage{microtype}           % Improved kerning and line breaks

% === Layout ===
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\setstretch{1.08}                % Slightly more than single spacing
\usepackage{parskip}             % Paragraph spacing instead of indent
\setlength{\parskip}{0.5\baselineskip}
\setlength{\parindent}{0pt}

% === Mathematics ===
\usepackage{amsmath,amssymb,amsthm}

% === Tables and Figures ===
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{caption}
\usepackage{subcaption}
\captionsetup{
  font={small},
  labelfont={bf},
  labelsep=period,
  skip=6pt
}

% === Algorithms ===
\usepackage{algorithm}
\usepackage{algorithmic}

% === Colors and Links ===
\usepackage{xcolor}
\definecolor{linkblue}{RGB}{0,51,153}
\definecolor{citegreen}{RGB}{0,102,51}
\usepackage[
  colorlinks=true,
  linkcolor=linkblue,
  citecolor=citegreen,
  urlcolor=linkblue,
  bookmarks=true,
  bookmarksnumbered=true,
  pdfstartview=FitH,
  pdftitle={S(M,T): Scoring AI Endpoints and the Measurement Gap in LLM Routing},
  pdfauthor={Pranav Lakherwal}
]{hyperref}

% === Bibliography ===
\usepackage{natbib}
\bibliographystyle{plainnat}

% === Headers and Footers ===
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small\textit{S(M,T): Scoring AI Endpoints}}
\fancyhead[R]{\small\thepage}
\renewcommand{\headrulewidth}{0.4pt}
\fancypagestyle{plain}{%
  \fancyhf{}
  \fancyfoot[C]{\small\thepage}
  \renewcommand{\headrulewidth}{0pt}
}

% === Section Formatting ===
\usepackage{titlesec}
\titleformat{\section}{\large\bfseries}{\thesection}{1em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{1em}{}
\titleformat{\subsubsection}{\normalsize\itshape}{\thesubsubsection}{1em}{}
\titlespacing*{\section}{0pt}{1.5\baselineskip}{0.75\baselineskip}
\titlespacing*{\subsection}{0pt}{1\baselineskip}{0.5\baselineskip}
\titlespacing*{\subsubsection}{0pt}{0.75\baselineskip}{0.25\baselineskip}

% === Theorem Environments ===
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}

% === Custom Commands ===
\newcommand{\smt}{S(M,T)}
\newcommand{\phii}{\phi_i}
\newcommand{\betavec}{\boldsymbol{\beta}}
\newcommand{\wvec}{\mathbf{w}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Prob}{\mathbb{P}}

% =====================================================================
\begin{document}

% === Title Block ===
\begin{center}
{\LARGE\bfseries S(M,T): Scoring AI Endpoints\\[4pt]
and the Measurement Gap in LLM Routing\par}

\vspace{14pt}

{\large Pranav Lakherwal}\\[4pt]
{\normalsize\texttt{pranav@pranavlakherwal.com}}

\vspace{10pt}

{\normalsize March 2026}\\[4pt]
{\small\textit{v1.1. Restructured for clarity and conciseness.\\
Convergence proofs moved to appendix. Added design reasoning,\\
self-critique, and Goodhart framing. See CHANGELOG.md for details.\\
Previous version: \texttt{github.com/pranavlakherwal/smt-router/releases/tag/v1.0}}}
\end{center}

\vspace{10pt}

% === Abstract ===
\begin{abstract}
\noindent
Routing AI queries to the right endpoint is becoming critical infrastructure as organizations spend billions on heterogeneous AI systems (LLMs, agents, scripts, tools). However, no existing framework scores these diverse endpoint types through a single equation, and no prior work documents why hand-designed scoring functions fail systematically.

We propose S(M,T), a scoring framework that multiplies three terms: binary gates enforcing hard constraints, a weighted compatibility function, and a task-conditional cost penalty. Unlike existing routers, the gate layer encodes type-specific constraints (scripts have no context window, agents have no token cost), enabling a single equation to route across heterogeneous endpoint pools.

We attempted to compute the compatibility functions from 2.76M public benchmark records. Fourteen designs across four research cycles all failed. The reasons are structural: metrics use incomparable scales, benchmark performance does not transfer to deployment, and complex properties collapse when reduced to single numbers. We call this the measurement gap.

Replacing hand-designed functions with 16 learned bilinear forms (3.15M parameters, trained on 740K routing samples) achieves 83.63\% accuracy on RouterBench and AUC 0.8006 on RouteLLM, retaining 94.35\% of best-model quality at half the cost. The model generalizes across both benchmarks without benchmark-specific tuning. A 4.6$\times$ scaling experiment confirms that data quality, not model capacity, is the binding constraint.

This enables a path toward general-purpose AI routing: one equation that scores any endpoint against any task, where the architect chooses the routing framework that fits their problem. Code: \texttt{github.com/pranavlakherwal/smt-router}.
\end{abstract}

\vspace{6pt}
\hrule
\vspace{12pt}

% === Body ===
\input{sections/01_intro}
\input{sections/02_equation}
\input{sections/03_measurement_gap}
\input{sections/04_learned_phi}
\input{sections/05_scaling}
\input{sections/06_theory}
\input{sections/07_related_work}
\input{sections/08_discussion}
\input{sections/09_conclusion}

% === Bibliography ===
\bibliography{bibliography}

% === Appendices ===
\appendix
\input{sections/appendix_convergence}
\input{sections/appendix_adversarial}

\end{document}
